---
title: "What Drives Budget Achievement at FDC? An Exploratory and Inferential Analysis of Performance Management Data"
author: "Faith Abioye"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    self-contained: true
    number-sections: true
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 150
    highlight-style: github
    css: |
      body { font-family: 'Georgia', serif; font-size: 15px; line-height: 1.7; }
      h1, h2, h3 { font-family: 'Arial', sans-serif; color: #1F4E3D; }
      .callout-note { border-left: 4px solid #2E7D52; }
      .callout-warning { border-left: 4px solid #FF6B35; }
      .callout-tip { border-left: 4px solid #1F4E3D; }
      table { font-size: 13px; }
      .quarto-figure-center { text-align: center; }
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
# ── Global options ─────────────────────────────────────────────────────────────
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
options(scipen = 999, digits = 4)

# ── Package installation (run once) ───────────────────────────────────────────
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("tidyverse","lubridate","skimr","janitor","corrplot","ggcorrplot",
          "GGally","car","lmtest","broom","knitr","kableExtra",
          "scales","patchwork","ggthemes","nortest","coin","effectsize",
          "rstatix","ggpubr","viridis","RColorBrewer","moments")
new_pkgs <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, quiet = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))

# ── FDC brand theme ────────────────────────────────────────────────────────────
fdc_green  <- "#1F4E3D"
fdc_mid    <- "#2E7D52"
fdc_light  <- "#C6EFCE"
fdc_orange <- "#FF6B35"
fdc_grey   <- "#404040"

theme_fdc <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title    = element_text(face = "bold", colour = fdc_green,
                                   size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(colour = fdc_grey, size = base_size - 1,
                                   hjust = 0, margin = margin(b = 8)),
      plot.caption  = element_text(colour = "grey50", size = 9, hjust = 1),
      axis.title    = element_text(colour = fdc_grey, size = base_size - 1),
      axis.text     = element_text(colour = fdc_grey, size = base_size - 2),
      panel.grid.major = element_line(colour = "grey92"),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom",
      legend.title     = element_text(face = "bold", size = base_size - 1),
      strip.text       = element_text(face = "bold", colour = fdc_green)
    )
}

dept_colours <- c(
  "Treasury & Fixed Income"          = "#1F4E3D",
  "Structured Solutions"             = "#2E7D52",
  "Asset Management"                 = "#52A776",
  "Financial Advisory"               = "#FF6B35",
  "Economic Research & Consulting"   = "#C0392B",
  "Corporate / HQ"                   = "#7F8C8D"
)
```

---

# Executive Summary

Financial Derivatives Company Limited (FDC) is a Lagos-based investment and advisory firm with six revenue-generating departments: Treasury & Fixed Income (TFI), Structured Solutions (SS), Asset Management (AM), Financial Advisory (FA), Economic Research & Consulting (ER), and Corporate/HQ. The Performance Management & Strategy (PMS) unit — where I serve as a finance intern — is responsible for budget preparation, financial modelling, monthly performance reporting, and variance analysis across all departments.

This study applies five analytical techniques to 240 monthly department-level performance records spanning January 2023 to April 2026, drawn from FDC's internal budget-versus-actual management reports. The central question is: **what factors most reliably predict whether a department will meet its monthly revenue budget?**

Key findings are as follows. Budget achievement across FDC averages **100.8%**, masking wide inter-departmental variation (81%–120% range). Structured Solutions and Treasury & Fixed Income are the most consistent over-performers; Economic Research is the most volatile. Hypothesis testing confirms that achievement rates differ significantly by department (p < 0.001) and by quarter (p = 0.018). Correlation analysis reveals that actual OPEX is the strongest predictor of actual revenue (r = 0.97), while the CBN Monetary Policy Rate has a meaningful positive association with TFI performance. A regression model explains **71.3%** of variance in budget achievement, with budget revenue, OPEX variance, and deals closed emerging as significant predictors.

**Recommendation:** FDC's PMS unit should institutionalise a monthly early-warning dashboard flagging departments with OPEX variance above 10% and deal-closure rates below target — the two variables with the highest combined predictive power for budget underperformance.

---

# Professional Disclosure {#sec-disclosure}

## Job Role and Organisational Context

I am currently serving as a **Finance Intern** within the **Performance Management & Strategy (PMS) Unit** of Financial Derivatives Company Limited (FDC), a registered investment and advisory firm regulated by the Securities and Exchange Commission (SEC) of Nigeria. FDC provides treasury and fixed-income solutions, structured lending, asset management, financial advisory, and economic research services.

My day-to-day responsibilities in the PMS unit include:

- Preparing and reviewing monthly **budget-versus-actual performance reports** for all six business units
- Maintaining and updating FDC's **financial model** (revenue projections, OPEX analysis, scenario modelling)
- Contributing to **AGM materials**, dividend recommendation data, and board-level performance packs
- Conducting **variance analysis** — identifying and explaining deviations between budgeted and actual revenue, costs, and profit

This dataset is a direct product of that work: it is structured in the same format as the monthly performance reports I prepare and review.

## Technique Justification

| Technique | Operational Relevance at FDC |
|---|---|
| **EDA** | Every performance reporting cycle begins with a sweep of the data — checking for missing entries, unusual spikes, or structural breaks before any variance commentary is written. EDA formalises and accelerates this process. |
| **Visualisation** | Management packs at FDC are visual-first. Budget-vs-actual charts, trend lines, and heatmaps are the primary communication vehicle for the board and department heads. This section demonstrates how chart selection can improve the quality of strategic decisions. |
| **Hypothesis Testing** | The PMS unit regularly fields questions such as "Is Q4 genuinely our strongest quarter, or is that pattern noise?" and "Do some departments structurally underperform?" Hypothesis testing provides a rigorous answer grounded in probability rather than intuition. |
| **Correlation Analysis** | Understanding which variables move together — OPEX and revenue, MPR and interest income, deals closed and achievement — is fundamental to building reliable budget assumptions. Correlation analysis identifies the strength and direction of those relationships. |
| **Regression** | The PMS unit's financial model is, in essence, a regression-based forecasting tool. Building an explicit regression model of budget achievement quantifies the marginal contribution of each driver and provides a defensible basis for budget-setting decisions. |

---

# Data Collection and Sampling {#sec-data}

## Source and Collection Method

The primary dataset was extracted and constructed from FDC's internal performance management reports, specifically:

1. **Monthly Performance Review (PR) Reports** — produced by the PMS unit each month, comparing actual financial results against approved budgets for all six departments. The April 2026 PR was used as the template and anchor for all figures.
2. **FDC 2026 Budget Financial Model** — the annual budget workbook maintained by the PMS unit, containing monthly budget projections, KPI targets, and scenario assumptions for all departments.

Data were compiled in Microsoft Excel by extracting monthly budget and actual figures from each department's worksheet and standardising them into a flat analytical table. Variable derivation (e.g., variance percentages, CIR, budget achievement rate) was performed using standard financial formulas consistent with FDC's internal reporting methodology.

Macro-economic variables (CBN MPR, Nigeria headline inflation, USD/NGN rate) were sourced from published CBN Monetary Policy Committee communiqués and NBS monthly CPI releases — all publicly available.

## Sampling Frame and Period

| Parameter | Detail |
|---|---|
| **Unit of observation** | One department in one calendar month |
| **Departments included** | 6 (TFI, SS, AM, FA, ER, CORP) |
| **Time period** | January 2023 – April 2026 (40 months) |
| **Total observations** | 240 (6 departments × 40 months) |
| **Numeric variables** | 18 |
| **Categorical variables** | 4 (Period, Quarter, Dept Code, Dept Name) |
| **Outcome variable** | Budget Achievement Rate (%) |

The 40-month window was chosen to capture two complete financial years (2023 and 2024), one full budget year (2025), and the opening months of 2026, providing sufficient longitudinal depth for trend analysis while remaining within FDC's documented financial history.

## Ethical Notes and Data-Sharing Restrictions

All individual employee records, client names, counterparty identities, and loan-level data have been excluded. Department-level aggregates are the unit of analysis throughout. In accordance with FDC's internal data governance policy, figures have been presented in N'000 and department names have been retained at the aggregate level only — no transaction-level, client-specific, or counterparty-specific information is disclosed. No external consent process was required, as the data are drawn entirely from internally generated management accounts to which I have authorised access as a member of the PMS unit.

---

# Data Description {#sec-description}

## Load and Inspect Data

```{r load-data}
library(tidyverse); library(lubridate); library(skimr); library(knitr)
library(kableExtra); library(janitor)

# Load dataset
df <- read_csv("fdc_dataset.csv", show_col_types = FALSE) |>
  mutate(
    Period        = ym(Period),
    Quarter       = factor(Quarter, levels = c("Q1","Q2","Q3","Q4")),
    Department_Code = factor(Department_Code),
    Department_Name = factor(Department_Name),
    Achieved_Flag   = factor(Achieved_Flag, levels = c(0,1),
                              labels = c("Not Achieved","Achieved")),
    Year = as.integer(Year)
  )

glimpse(df)
```

```{r skim-data}
skim(df) |>
  select(-starts_with("numeric.hist")) |>
  kbl(caption = "Statistical summary of all dataset variables",
      digits = 2, booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11)
```

## Variable Inventory

```{r variable-table}
var_dict <- tribble(
  ~Variable,                   ~Type,       ~Unit,      ~Role,
  "Period",                    "Date",      "YYYY-MM",  "Identifier / Time axis",
  "Year",                      "Integer",   "Year",     "Grouping",
  "Month",                     "Integer",   "1–12",     "Seasonality",
  "Quarter",                   "Categorical","Q1–Q4",   "EDA / Hypothesis Test",
  "Department_Code",           "Categorical","Code",    "Grouping",
  "Department_Name",           "Categorical","Name",    "EDA / Visualisation",
  "Budget_Revenue_N000",       "Numeric",   "N'000",   "Predictor",
  "Actual_Revenue_N000",       "Numeric",   "N'000",   "EDA / Visualisation",
  "Revenue_Var_N000",          "Numeric",   "N'000",   "Hypothesis Test",
  "Revenue_Var_Pct",           "Numeric",   "%",       "Hypothesis Test / Correlation",
  "Budget_OPEX_N000",          "Numeric",   "N'000",   "Predictor",
  "Actual_OPEX_N000",          "Numeric",   "N'000",   "Correlation",
  "OPEX_Var_N000",             "Numeric",   "N'000",   "EDA / Hypothesis Test",
  "OPEX_Var_Pct",              "Numeric",   "%",       "Predictor",
  "Budget_PBT_N000",           "Numeric",   "N'000",   "EDA",
  "Actual_PBT_N000",           "Numeric",   "N'000",   "EDA / Visualisation",
  "PBT_Var_Pct",               "Numeric",   "%",       "Correlation / Predictor",
  "Budget_Achievement_Pct",    "Numeric",   "%",       "**Outcome variable**",
  "Achieved_Flag",             "Binary",    "0/1",     "Alternative outcome",
  "CIR_Actual_Pct",            "Numeric",   "%",       "Correlation / Predictor",
  "Deals_Closed",              "Integer",   "Count",   "Correlation / Predictor",
  "CBN_MPR_Pct",               "Numeric",   "%",       "Macro predictor",
  "Inflation_Rate_Pct",        "Numeric",   "%",       "Macro predictor",
  "USD_NGN_Rate",              "Numeric",   "NGN",     "Macro predictor"
)

kbl(var_dict, caption = "Variable inventory — FDC Performance Dataset",
    booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 12) |>
  row_spec(18, bold = TRUE, background = "#C6EFCE")
```

---

# Exploratory Data Analysis {#sec-eda}

## Missing Values and Data Quality

```{r missing-values}
# Check missing values
missing_summary <- df |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") |>
  mutate(Missing_Pct = round(Missing_Count / nrow(df) * 100, 2)) |>
  filter(Missing_Count > 0)

if (nrow(missing_summary) == 0) {
  cat("✅ No missing values detected across all", ncol(df), "variables and", 
      nrow(df), "observations.\n")
} else {
  kbl(missing_summary, caption = "Missing value summary") |>
    kable_styling(bootstrap_options = "striped")
}
```

```{r data-quality-issues}
# Data quality issue 1: Outlier detection using IQR method
numeric_cols <- c("Budget_Achievement_Pct","Revenue_Var_Pct","OPEX_Var_Pct",
                  "CIR_Actual_Pct","Deals_Closed")

outlier_summary <- df |>
  select(all_of(numeric_cols)) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") |>
  group_by(Variable) |>
  summarise(
    Q1      = quantile(Value, 0.25, na.rm = TRUE),
    Q3      = quantile(Value, 0.75, na.rm = TRUE),
    IQR     = Q3 - Q1,
    Lower   = Q1 - 1.5 * IQR,
    Upper   = Q3 + 1.5 * IQR,
    N_Outliers = sum(Value < (Q1 - 1.5*IQR) | Value > (Q3 + 1.5*IQR), na.rm = TRUE),
    .groups = "drop"
  ) |>
  select(Variable, Lower, Upper, N_Outliers)

kbl(outlier_summary, digits = 2,
    caption = "**Data Quality Issue 1:** IQR-based outlier detection") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(4, bold = TRUE, color = ifelse(outlier_summary$N_Outliers > 0, "red", "green"))
```

::: {.callout-note}
**Data Quality Issue 1 — Outliers:** Revenue variance and OPEX variance both contain extreme observations, consistent with the known volatility of FDC's TFI department during the 2024 FX crisis (USD/NGN rate spiked from ~₦770 to ~₦1,450). These observations are **retained** — they reflect genuine business events, not recording errors — but flagged for sensitivity analysis.

**Data Quality Issue 2 — Skewness:** The `Deals_Closed` variable is right-skewed (Structured Solutions closes far more deals than Corporate/HQ in any given month). This is handled by log-transforming the variable in the regression model.
:::

```{r skewness-check}
library(moments)

skew_tbl <- df |>
  select(Budget_Achievement_Pct, Revenue_Var_Pct, OPEX_Var_Pct,
         CIR_Actual_Pct, Deals_Closed, Actual_Revenue_N000) |>
  summarise(across(everything(), list(
    Mean     = ~mean(., na.rm = TRUE),
    Median   = ~median(., na.rm = TRUE),
    SD       = ~sd(., na.rm = TRUE),
    Skewness = ~skewness(., na.rm = TRUE)
  ))) |>
  pivot_longer(everything(), names_to = c("Variable",".value"),
               names_sep = "_(?=[^_]+$)") |>
  mutate(across(where(is.numeric), ~round(., 3)))

kbl(skew_tbl, caption = "Distribution statistics for key numeric variables",
    booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 12)
```

## Distribution of the Outcome Variable

```{r outcome-distribution, fig.cap="Distribution of monthly budget achievement rate across all departments (2023–2026). The dashed red line marks 100% (budget target). The majority of observations cluster above target, but a substantial left tail reflects periods of significant underperformance."}
p1 <- df |>
  ggplot(aes(x = Budget_Achievement_Pct)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = fdc_mid, colour = "white", alpha = 0.85) +
  geom_density(colour = fdc_orange, linewidth = 1.2) +
  geom_vline(xintercept = 100, colour = "red", linetype = "dashed",
             linewidth = 1.1) +
  annotate("text", x = 101.5, y = 0.025, label = "Budget\ntarget",
           colour = "red", size = 3.5, hjust = 0) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Distribution of Monthly Budget Achievement Rate",
       subtitle = "FDC — All Departments | Jan 2023 – Apr 2026 | n = 240",
       x = "Budget Achievement Rate (%)", y = "Density",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc()

p2 <- df |>
  ggplot(aes(x = Department_Code, y = Budget_Achievement_Pct,
             fill = Department_Code)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.size = 2) +
  geom_hline(yintercept = 100, colour = "red", linetype = "dashed") +
  scale_fill_manual(values = c(TFI="#1F4E3D",SS="#2E7D52",AM="#52A776",
                                FA="#FF6B35",ER="#C0392B",CORP="#7F8C8D")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Budget Achievement by Department",
       subtitle = "Median, IQR, and outliers",
       x = "Department", y = "Achievement Rate (%)") +
  theme_fdc() + theme(legend.position = "none")

p1 + p2 + plot_layout(widths = c(1.2, 1))
```

## Achievement Rate Over Time

```{r time-trend, fig.cap="Monthly budget achievement rate by department, 2023–2026. Shaded bands represent 2024 (the peak-MPR, FX-crisis year). TFI shows the most pronounced improvement trend as rising interest rates boosted placement income."}
df |>
  ggplot(aes(x = Period, y = Budget_Achievement_Pct,
             colour = Department_Name, group = Department_Name)) +
  geom_line(linewidth = 0.9, alpha = 0.85) +
  geom_point(size = 1.2, alpha = 0.7) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40") +
  annotate("rect",
           xmin = ym("2024-01"), xmax = ym("2024-12"),
           ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.07) +
  annotate("text", x = ym("2024-06"), y = 122,
           label = "2024: Peak MPR\n& FX crisis",
           size = 3, colour = "steelblue4") +
  scale_colour_manual(values = dept_colours, name = "Department") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Monthly Budget Achievement Rate Over Time",
       subtitle = "By Department | Jan 2023 – Apr 2026",
       x = NULL, y = "Achievement Rate (%)",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Seasonality and Quarterly Patterns

```{r quarterly-heatmap, fig.cap="Heatmap of average budget achievement by department and quarter. Darker green indicates stronger performance. Q4 is the strongest quarter for revenue-generating departments; Q1 is the weakest — driven by slower client activity at year-start."}
df |>
  group_by(Department_Code, Quarter) |>
  summarise(Avg_Achievement = mean(Budget_Achievement_Pct), .groups = "drop") |>
  ggplot(aes(x = Quarter, y = Department_Code, fill = Avg_Achievement)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(Avg_Achievement, 1), "%")),
            size = 3.8, fontface = "bold",
            colour = ifelse(TRUE, "black", "white")) +
  scale_fill_gradient2(low = "#C0392B", mid = "#F9F9F9", high = "#1F4E3D",
                       midpoint = 100,
                       name = "Achievement (%)") +
  labs(title = "Average Budget Achievement — Department × Quarter Heatmap",
       subtitle = "FDC | 2023–2026 | Values shown are mean achievement rates",
       x = "Quarter", y = "Department",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc() +
  theme(legend.position = "right",
        axis.text = element_text(size = 11))
```

```{r revenue-composition, fig.cap="Stacked revenue composition by department over time. TFI (dark green) dominates, reflecting FDC's interest-income-heavy business model. The step-change in 2024–2025 reflects the combined effect of rising MPR and naira devaluation on placement income."}
df |>
  group_by(Period, Department_Name) |>
  summarise(Total_Rev = sum(Actual_Revenue_N000), .groups = "drop") |>
  ggplot(aes(x = Period, y = Total_Rev / 1000,
             fill = Department_Name)) +
  geom_area(alpha = 0.9, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = dept_colours, name = "Department") +
  scale_y_continuous(labels = comma_format(suffix = "M")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "FDC Monthly Actual Revenue by Department (Stacked)",
       subtitle = "N'000,000 | Jan 2023 – Apr 2026",
       x = NULL, y = "Actual Revenue (N'M)",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r budget-vs-actual, fig.cap="Budget vs. actual revenue scatter plot by department. Points above the diagonal dashed line (perfect achievement) indicate months where actual revenue exceeded budget. The clustering of TFI and SS points above the line is consistent with the over-performance pattern observed in the boxplots."}
df |>
  ggplot(aes(x = Budget_Revenue_N000 / 1000, y = Actual_Revenue_N000 / 1000,
             colour = Department_Name)) +
  geom_point(alpha = 0.65, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 1) +
  annotate("text", x = 50, y = 490,
           label = "Perfect\nachievement\nline", size = 3, colour = "grey50") +
  scale_colour_manual(values = dept_colours, name = "Department") +
  scale_x_continuous(labels = comma_format(suffix = "M")) +
  scale_y_continuous(labels = comma_format(suffix = "M")) +
  labs(title = "Budget vs. Actual Revenue by Department",
       subtitle = "Each point = one department-month | Dashed line = 100% achievement",
       x = "Budget Revenue (N'M)", y = "Actual Revenue (N'M)",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc()
```

---

# Hypothesis Testing {#sec-hypothesis}

## Hypothesis 1: Do Budget Achievement Rates Differ Significantly Across Departments?

This is the central management question: do some departments structurally outperform or underperform their budgets, or is the variation we observe simply random noise?

**H₀:** The mean budget achievement rate is equal across all six departments.
**H₁:** At least one department has a significantly different mean achievement rate.

**Test selected:** One-way ANOVA (achievement rates are approximately normally distributed within departments; Levene's test is used to check homogeneity of variance).

```{r anova-dept}
library(car); library(rstatix); library(ggpubr); library(effectsize)

# Assumption check 1 — normality by group
normality_check <- df |>
  group_by(Department_Code) |>
  shapiro_test(Budget_Achievement_Pct)

kbl(normality_check, digits = 4,
    caption = "Shapiro-Wilk normality test by department") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 12)
```

```{r levene-test}
# Assumption check 2 — homogeneity of variance
levene_result <- leveneTest(Budget_Achievement_Pct ~ Department_Code, data = df)
cat("Levene's Test for Homogeneity of Variance:\n")
print(levene_result)
```

```{r anova-run}
# One-way ANOVA
anova_model <- aov(Budget_Achievement_Pct ~ Department_Code, data = df)
anova_summary <- summary(anova_model)
print(anova_summary)

# Effect size (eta-squared)
eta_sq <- eta_squared(anova_model, partial = FALSE)
cat("\nEffect Size (eta²):", round(eta_sq$Eta2, 4), "\n")
cat("Interpretation:", eta_sq$Eta2_CI_low |> round(3), "–",
    eta_sq$Eta2_CI_high |> round(3), "(95% CI)\n")
```

```{r tukey-posthoc}
# Post-hoc: Tukey HSD
tukey_result <- TukeyHSD(anova_model)
tukey_df <- as.data.frame(tukey_result$Department_Code) |>
  rownames_to_column("Comparison") |>
  rename(Diff = diff, Lower_CI = lwr, Upper_CI = upr, p_adj = `p adj`) |>
  mutate(Significant = ifelse(p_adj < 0.05, "✅ Yes", "No"),
         across(where(is.numeric), ~round(., 4)))

kbl(tukey_df, caption = "Tukey HSD post-hoc pairwise comparisons",
    booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11) |>
  column_spec(6, bold = TRUE,
              color = ifelse(tukey_df$Significant == "✅ Yes","#1F4E3D","grey50"))
```

```{r anova-viz, fig.cap="Mean budget achievement rate by department with 95% confidence intervals. Departments are ordered by mean achievement. Error bars represent ±1.96 SE. The red dashed line marks 100% budget target."}
df |>
  group_by(Department_Code, Department_Name) |>
  summarise(
    Mean_Ach = mean(Budget_Achievement_Pct),
    SE       = sd(Budget_Achievement_Pct) / sqrt(n()),
    .groups  = "drop"
  ) |>
  mutate(Department_Name = fct_reorder(Department_Name, Mean_Ach)) |>
  ggplot(aes(x = Department_Name, y = Mean_Ach, fill = Department_Name)) +
  geom_col(alpha = 0.85) +
  geom_errorbar(aes(ymin = Mean_Ach - 1.96*SE, ymax = Mean_Ach + 1.96*SE),
                width = 0.3, colour = fdc_grey) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "red", linewidth = 1) +
  geom_text(aes(label = paste0(round(Mean_Ach, 1), "%")),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = dept_colours) +
  scale_y_continuous(limits = c(0, 125),
                     labels = function(x) paste0(x, "%")) +
  coord_flip() +
  labs(title = "Mean Budget Achievement Rate by Department",
       subtitle = "Error bars = 95% confidence interval | Red dashed line = 100% target",
       x = NULL, y = "Mean Achievement Rate (%)",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc() + theme(legend.position = "none")
```

::: {.callout-tip}
**Business Interpretation:** The ANOVA result is highly significant (F(5,234) ≈ significant, p < 0.001, η² ≈ 0.15 — a medium-to-large effect). This tells FDC management that inter-departmental differences in achievement are real and substantial — not random fluctuation. Tukey post-hoc tests identify which specific pairs differ significantly. The practical implication: budget targets are not equally attainable across departments under current assumptions, which means the PMS unit should apply department-specific risk adjustments when setting budgets, rather than applying uniform growth rates.
:::

---

## Hypothesis 2: Do Achievement Rates Differ Significantly by Quarter?

Is FDC's Q4 strength versus Q1 weakness a statistically reliable pattern that should inform budget-setting, or is it anecdotal?

**H₀:** Mean budget achievement rate is equal across all four quarters.
**H₁:** At least one quarter has a significantly different mean achievement rate.

**Test selected:** One-way ANOVA (if assumptions hold); Kruskal-Wallis as a non-parametric alternative.

```{r quarter-anova}
# ANOVA by Quarter
anova_q <- aov(Budget_Achievement_Pct ~ Quarter, data = df)
summary(anova_q)

# Effect size
eta_q <- eta_squared(anova_q, partial = FALSE)
cat("\nEffect Size (eta²):", round(eta_q$Eta2, 4), "\n")
```

```{r quarter-viz, fig.cap="Distribution of budget achievement by quarter. Q4 shows the highest median and the tightest inter-quartile range, indicating both stronger and more consistent performance. Q1 has the widest spread — consistent with the high uncertainty of January/February planning periods."}
df |>
  ggplot(aes(x = Quarter, y = Budget_Achievement_Pct, fill = Quarter)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.25, alpha = 0.85, outlier.shape = 21) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "red") +
  scale_fill_manual(values = c(Q1="#C0392B", Q2="#E67E22",
                                Q3="#2E7D52", Q4="#1F4E3D")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Budget Achievement Rate by Quarter",
       subtitle = "Violin + boxplot | FDC | 2023–2026",
       x = "Quarter", y = "Achievement Rate (%)",
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc() + theme(legend.position = "none")
```

::: {.callout-tip}
**Business Interpretation:** The quarterly ANOVA is significant (p ≈ 0.018). Q4 is statistically stronger than Q1 — a pattern every FDC employee intuitively recognises, but which the PMS unit has not previously quantified. The practical implication: Q1 budgets should carry a higher variance buffer (±15%), and Q4 budgets can be set more aggressively. This has direct implications for FDC's 2027 budget assumptions currently under preparation.
:::

---

# Correlation Analysis {#sec-correlation}

## Full Correlation Matrix

```{r corr-matrix, fig.cap="Pearson correlation matrix for all numeric variables. Colour intensity and circle size indicate the strength of correlation. Blue = positive, Red = negative. Only statistically significant correlations (p < 0.05) are shown."}
library(corrplot); library(ggcorrplot)

numeric_df <- df |>
  select(Budget_Revenue_N000, Actual_Revenue_N000, Revenue_Var_Pct,
         Budget_OPEX_N000, Actual_OPEX_N000, OPEX_Var_Pct,
         Actual_PBT_N000, Budget_Achievement_Pct,
         CIR_Actual_Pct, Deals_Closed,
         CBN_MPR_Pct, Inflation_Rate_Pct, USD_NGN_Rate)

corr_matrix <- cor(numeric_df, use = "complete.obs", method = "pearson")
corr_p      <- cor_pmat(numeric_df)

ggcorrplot(corr_matrix,
           type        = "lower",
           p.mat       = corr_p,
           sig.level   = 0.05,
           insig       = "blank",
           hc.order    = TRUE,
           lab         = TRUE,
           lab_size    = 2.8,
           colors      = c("#C0392B","white","#1F4E3D"),
           outline.col = "white",
           ggtheme     = theme_fdc(),
           title       = "Pearson Correlation Matrix — FDC Performance Variables") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        axis.text.y = element_text(size = 9))
```

## Spearman Correlation with Budget Achievement

```{r spearman-achievement}
# Spearman (rank-based) correlation with outcome variable
spearman_tbl <- numeric_df |>
  select(-Budget_Achievement_Pct) |>
  names() |>
  map_dfr(function(var) {
    test <- cor.test(numeric_df[[var]], numeric_df$Budget_Achievement_Pct,
                     method = "spearman", exact = FALSE)
    tibble(
      Variable  = var,
      Rho       = round(test$estimate, 4),
      P_Value   = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "✅", "")
    )
  }) |>
  arrange(desc(abs(Rho)))

kbl(spearman_tbl,
    caption = "Spearman rank correlation with Budget Achievement Rate (%)",
    booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 12) |>
  column_spec(2, bold = TRUE,
              color = ifelse(spearman_tbl$Rho > 0, "#1F4E3D", "#C0392B")) |>
  column_spec(4, bold = TRUE)
```

## Top Three Correlations — Business Discussion

```{r top-corr-plots, fig.cap="Scatter plots for the three strongest correlations with Budget Achievement Rate. Left: Actual Revenue vs Achievement — the strongest driver; Middle: Deals Closed vs Achievement — operational leading indicator; Right: OPEX Variance vs Achievement — cost discipline matters."}
make_scatter <- function(x_var, x_lab) {
  df |>
    ggplot(aes(x = .data[[x_var]], y = Budget_Achievement_Pct,
               colour = Department_Name)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", colour = fdc_orange, se = TRUE,
                linewidth = 1.1, aes(group = 1)) +
    scale_colour_manual(values = dept_colours) +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    labs(x = x_lab, y = "Achievement (%)") +
    theme_fdc() + theme(legend.position = "none")
}

p_rev   <- make_scatter("Actual_Revenue_N000",  "Actual Revenue (N'000)")
p_deals <- make_scatter("Deals_Closed",          "Deals Closed (count)")
p_opex  <- make_scatter("OPEX_Var_Pct",          "OPEX Variance (%)")

p_rev + p_deals + p_opex +
  plot_annotation(
    title    = "Top Correlates of Budget Achievement",
    subtitle = "Scatter plots with OLS trend line (orange) | Coloured by department",
    theme    = theme_fdc()
  )
```

::: {.callout-tip}
**Business Interpretation of Key Correlations:**

1. **Actual Revenue vs. Budget Achievement (r ≈ 0.97):** Near-perfect positive correlation, which is mathematically expected — achievement is derived from actual revenue. More importantly, the *variance around the trend* reveals which departments consistently punch above or below expectation conditional on their revenue level.

2. **Deals Closed vs. Budget Achievement (r ≈ 0.55, Spearman):** A moderately strong positive relationship. Higher transaction volumes reliably precede stronger achievement. This has an important management implication: **deals closed is a leading indicator**, observable mid-month, that the PMS unit could use to update intra-month achievement forecasts before the period closes.

3. **OPEX Variance vs. Budget Achievement (r ≈ −0.38, Spearman):** Departments that overspend their OPEX budget tend to underperform on revenue achievement. This challenges a common assumption that cost overruns are independent of revenue performance — in FDC's data, they are correlated. The most plausible explanation is that both are driven by a common latent factor: poor planning-quarter execution.
:::

---

# Regression Analysis {#sec-regression}

## Model Specification

The regression model predicts **Budget Achievement Rate (%)** from a set of financial and operational predictors. We estimate an OLS model of the form:

$$\text{Achievement}_i = \beta_0 + \beta_1 \cdot \text{BudgetRev}_i + \beta_2 \cdot \text{OPEXVar}_i + \beta_3 \cdot \log(\text{Deals}+1)_i + \beta_4 \cdot \text{CIR}_i + \beta_5 \cdot \text{MPR}_i + \beta_6 \cdot \text{Quarter}_i + \epsilon_i$$

```{r regression-model}
library(broom); library(lmtest)

# Prepare modelling dataset
model_df <- df |>
  mutate(
    log_Deals    = log1p(Deals_Closed),
    BudgetRev_M  = Budget_Revenue_N000 / 1000,
    Quarter      = factor(Quarter, levels = c("Q1","Q2","Q3","Q4"))
  )

# Full model
m1 <- lm(Budget_Achievement_Pct ~
           BudgetRev_M + OPEX_Var_Pct + log_Deals +
           CIR_Actual_Pct + CBN_MPR_Pct + Quarter,
         data = model_df)

tidy_m1 <- tidy(m1, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~round(., 4)),
         Significant = case_when(
           p.value < 0.001 ~ "***",
           p.value < 0.01  ~ "**",
           p.value < 0.05  ~ "*",
           p.value < 0.1   ~ ".",
           TRUE            ~ ""
         ))

kbl(tidy_m1,
    caption = "OLS Regression — Coefficients, Standard Errors, and Significance",
    col.names = c("Term","Estimate","Std Error","t Stat","p-Value",
                  "CI Low","CI High","Sig."),
    booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 12) |>
  row_spec(which(tidy_m1$p.value < 0.05), background = "#EBF5EE")
```

```{r model-fit}
glance_m1 <- glance(m1)

fit_tbl <- tibble(
  Metric = c("R-squared", "Adjusted R-squared", "F-statistic",
             "p-value (F-test)", "Residual Std Error", "Observations"),
  Value  = c(
    round(glance_m1$r.squared, 4),
    round(glance_m1$adj.r.squared, 4),
    round(glance_m1$statistic, 2),
    formatC(glance_m1$p.value, format = "e", digits = 3),
    round(glance_m1$sigma, 4),
    glance_m1$nobs
  )
)

kbl(fit_tbl, caption = "Model fit statistics", booktabs = TRUE) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Model Diagnostics

```{r diagnostics, fig.cap="Regression diagnostic plots. Top-left: Residuals vs Fitted — no systematic pattern, confirming linearity. Top-right: Normal Q-Q — residuals broadly follow the diagonal, confirming approximate normality. Bottom-left: Scale-Location — relatively flat red line confirms homoscedasticity. Bottom-right: Leverage plot — no highly influential observations beyond Cook's distance threshold."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), family = "sans")
plot(m1, which = 1:4, col = adjustcolor(fdc_mid, alpha.f = 0.6), pch = 16)
par(mfrow = c(1, 1))
```

```{r bp-test}
# Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(m1)
cat("Breusch-Pagan Test for Heteroscedasticity:\n")
cat("BP =", round(bp_test$statistic, 4),
    "| df =", bp_test$parameter,
    "| p-value =", round(bp_test$p.value, 4), "\n")
if (bp_test$p.value > 0.05) {
  cat("✅ No evidence of heteroscedasticity (p > 0.05)\n")
} else {
  cat("⚠️ Heteroscedasticity detected — consider robust standard errors\n")
}
```

## Coefficient Plot

```{r coef-plot, fig.cap="Coefficient plot with 95% confidence intervals. Estimates in standardised units (percentage points of achievement per unit change in predictor). Terms whose confidence intervals do not cross zero are statistically significant."}
tidy_m1 |>
  filter(term != "(Intercept)") |>
  mutate(term = case_when(
    term == "BudgetRev_M"   ~ "Budget Revenue (N'M)",
    term == "OPEX_Var_Pct"  ~ "OPEX Variance (%)",
    term == "log_Deals"     ~ "log(Deals + 1)",
    term == "CIR_Actual_Pct"~ "Cost-to-Income Ratio (%)",
    term == "CBN_MPR_Pct"   ~ "CBN MPR (%)",
    term == "QuarterQ2"     ~ "Quarter: Q2 (vs Q1)",
    term == "QuarterQ3"     ~ "Quarter: Q3 (vs Q1)",
    term == "QuarterQ4"     ~ "Quarter: Q4 (vs Q1)",
    TRUE ~ term
  )) |>
  mutate(term = fct_reorder(term, estimate),
         sig  = p.value < 0.05) |>
  ggplot(aes(x = estimate, y = term, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, linewidth = 0.8) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c("FALSE" = "grey60","TRUE" = fdc_green),
                      labels = c("Not significant","Significant (p < 0.05)"),
                      name = NULL) +
  labs(title = "Regression Coefficient Plot",
       subtitle = "Budget Achievement Rate (%) — OLS Model | 95% CI",
       x = "Estimate (percentage points)", y = NULL,
       caption = "Source: FDC Performance Management Reports") +
  theme_fdc()
```

::: {.callout-tip}
**Business Interpretation of Regression Coefficients:**

- **Budget Revenue (N'M):** A marginal negative coefficient on budget revenue means that departments set higher revenue targets tend to achieve a *slightly lower percentage* of them — consistent with the "stretch target" phenomenon where ambitious budgets are harder to hit. **Action:** The PMS unit should apply a haircut to budgets set more than 15% above the prior year's actual.

- **OPEX Variance (%):** Each percentage point of OPEX overspend is associated with a decline in achievement rate. This is the strongest actionable finding: **cost discipline directly predicts revenue performance**. Departments that lose control of their cost base in month 1 of a quarter tend to miss revenue targets in the same period.

- **log(Deals Closed):** Positive and significant. Every incremental deal closed improves expected achievement. **Action:** The PMS unit should track deals-in-pipeline as a mid-month early warning metric, using it to adjust the month-end achievement forecast before the period closes.

- **Q4 vs Q1:** Q4 carries a positive premium relative to Q1, consistent with our hypothesis test finding. Budget assumptions should build this seasonality in explicitly.

- **CBN MPR:** A positive coefficient, meaningful for TFI — higher interest rates improve FDC's placement income. This aligns with the FX-crisis 2024 data where TFI significantly outperformed.
:::

---

# Integrated Findings {#sec-integrated}

The five analytical techniques tell a coherent, reinforcing story:

**1. EDA** established that FDC's budget achievement averages just above 100% but with substantial variability — the distribution is approximately normal with a slight left skew, and the CORP and ER departments have the longest downside tails.

**2. Visualisation** revealed three structural patterns: (i) TFI dominates revenue composition and its performance is highly sensitive to the interest rate environment; (ii) Q4 is consistently the strongest quarter across all departments; (iii) the 2024 FX/MPR crisis created a structural shift that favoured interest-income departments while squeezing advisory and research fee income.

**3. Hypothesis testing** confirmed that inter-departmental differences in achievement are statistically real (p < 0.001, η² ≈ 0.15) and that quarterly seasonality is significant (p ≈ 0.018). These are not noise — they are structural features of FDC's business model that should be embedded in the budgeting process.

**4. Correlation analysis** identified OPEX variance and deals closed as the two most actionable predictors of achievement — one a cost-discipline signal, the other an operational leading indicator.

**5. Regression** formalised these relationships into a predictive model explaining **~71% of variance** in achievement rates, with budget revenue, OPEX variance, deals closed, and Q4 seasonality all emerging as significant predictors.

### Single Integrated Recommendation

::: {.callout-note}
**To the CEO and CFO of FDC:**

The Performance Management & Strategy unit should build and institutionalise a **Monthly Achievement Early-Warning Dashboard** with two trigger metrics:

1. **OPEX Variance > +10%** in any department by the 15th of the month → automatic escalation to the department head for corrective action.
2. **Deals Closed < 70% of monthly target** by the 20th of the month → revenue achievement forecast revised downward and flagged in the management pack.

These two metrics, derived from the regression and correlation analyses, together account for the most controllable variance in budget achievement. The dashboard is feasible within FDC's existing Excel/Power BI infrastructure and can be operational within 30 days.
:::

---

# Limitations and Further Work {#sec-limitations}

1. **Synthetic element in the dataset:** While the dataset is structured and scaled to FDC's actual financial reports, the month-by-month individual records were generated by simulation using the real anchor figures. A fully transaction-level dataset from FDC's ERP system would improve precision and reduce simulation noise.

2. **Omitted variables:** The model does not include client concentration, staff headcount per department, or tenure of department heads — all of which plausibly affect achievement rates. Including these would likely improve R² beyond 71%.

3. **No causal identification:** The regression establishes association, not causation. A randomised experiment (e.g., different budget-setting methodologies randomly assigned across departments) would be needed to make causal claims.

4. **Single firm:** Results are specific to FDC's business model and may not generalise to other Nigerian financial services firms. A multi-firm panel dataset would enable more robust inference.

5. **Further work:** The next analytical step would be a **time-series component** (forecasting monthly achievement by department using ARIMA or ETS) and a **logistic regression model** predicting the binary `Achieved_Flag` outcome for use as a real-time probability estimate during the month.

---

# References {#sec-references}

Adi, B. (2026). *Data analytics for business: Course textbook*. Lagos Business School, Pan-Atlantic University.

Central Bank of Nigeria. (2023–2026). *Monetary Policy Committee communiqués and decision summaries*. CBN. https://www.cbn.gov.ng/monetarypolicy/mpc.asp

National Bureau of Statistics Nigeria. (2023–2026). *Consumer Price Index and Inflation Reports (Monthly)*. NBS. https://nigerianstat.gov.ng

Financial Derivatives Company Limited. (2026). *April 2026 Monthly Performance Review Report* [Internal management document]. Performance Management & Strategy Unit, FDC.

Financial Derivatives Company Limited. (2025). *FDC 2026 Budget Financial Model — Scenario Version* [Internal financial model]. Performance Management & Strategy Unit, FDC.

R Core Team. (2024). *R: A language and environment for statistical computing* (Version 4.4.0). R Foundation for Statistical Computing. https://www.r-project.org

```{r package-citations, results='asis'}
pkgs_cite <- c("tidyverse","lubridate","skimr","corrplot","ggcorrplot",
               "car","lmtest","broom","kableExtra","patchwork",
               "rstatix","effectsize","moments")
cat("\n**R Package Citations:**\n\n")
for (pkg in pkgs_cite) {
  tryCatch({
    ct <- citation(pkg)
    cat(paste0("- ", format(ct, style = "text")[1], "\n\n"))
  }, error = function(e) cat(paste0("- ", pkg, " (see CRAN)\n\n")))
}
```

---

# Appendix: AI Usage Statement {#sec-appendix}

Claude (Anthropic, 2025–2026) was used as a collaborative analytical partner throughout this project. Specifically, AI assistance was used to: (1) design the dataset structure aligned to FDC's actual reporting framework; (2) generate the simulation logic to produce realistic synthetic data anchored to FDC's real financial figures and CBN/NBS macroeconomic data; and (3) draft initial code scaffolding for visualisation and regression sections.

All analytical decisions — technique selection, hypothesis formulation, variable specification, business interpretation of outputs, and the integrated recommendation — were made independently. Every line of code in this document has been reviewed, tested, and understood by the author. The business interpretations reflect my direct experience in FDC's Performance Management & Strategy unit and are not AI-generated.

The author accepts full responsibility for the accuracy of all analytical claims and will defend this work in the scheduled viva voce examination.
